---
title: "Polygenic risk scores predict hypertension"
editor_options:
  chunk_output_type: console
output:
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


**Libraries**


```{r}
setwd("/home/ivm/red/ht_prs")

packrat::on()
library(tidyverse)
library(survival)   # survival analysis
library(data.table) # fread() function
library(gridExtra)  # plots to grid
library(survminer)  # ggbased visualization and extra diagnostics
library(survRM2)    # restricted mean survival times
library(visdat)     # visualization of tibble and na's
library(forestplot) # forestplot

source('/finngen/red/ht_prs/articles-functions.R')
```


# Data

**Risk scores**

Calculated from UK Biobank (UKB) for Finngen Data Freeze 5 individuals using the PRS-CS18 pipeline with default parameters.


**Endpoints**

* I9_HYPTENS: Hypertension
* I9_CVD_HARD: Hard cardiovascular diseases
* I9_CHD: Major coronary heart disease event
* I9_STR: Ischaemic Stroke, excluding all haemorrhages.


**Copying and preprocessing data**
Endpoint and covariate files transfered and unzipped. Columns selected from endpoint file prior to import to r, because the original phenotype file is very large.


```{bash, eval=F}

#Endpoint file and covariate file copied and unzipped

cp /finngen/library-red/finngen_R5/phenotype_2.0/data/finngen_R5_V2_endpoint.gz /finngen/red/ht_prs/data/
cp /finngen/library-red/finngen_R5/phenotype_2.0/data/R5_cov_pheno_1.0.txt.gz /finngen/red/ht_prs/data/  

gzip -d /finngen/red/ht_prs/data/finngen_R5_V2_endpoint.gz
gzip -d /finngen/red/ht_prs/data/R5_cov_pheno_1.0.txt.gz

#Self written perl-script to extract columns from phenotype file
perl select_columns.pl data/finngen_R5_V2_endpoint data/finngen_R5_V2_HTCVD FINNGENID I9_HYPTENS I9_HYPTENS_AGE I9_CVD_HARD I9_CVD_HARD_AGE I9_CHD I9_CHD_AGE I9_STR I9_STR_AGE FU_END_AGE

#This could be loaded directly to R, but now loading will be faster...
perl select_columns.pl data/R5_cov_pheno_1.0.txt data/R5_cov_pheno_selected FINNGENID batch BL_YEAR cohort SEX_IMPUTED PC1 PC2 PC3 PC4 PC5 PC6 PC7 PC8 PC9 PC10

```


**Loading data**

```{r}

endpoints <- fread("/finngen/red/ht_prs/data/finngen_R5_V2_HTCVD", sep = "\t") 

sbp_prs <- fread("/finngen/red/ukb-bp-prs/ukb-sbp-both.profile") %>%
  select(IID, SCORE) %>%
  rename(SBP_SCORE=SCORE)

dbp_prs <- fread("/finngen/red/ukb-bp-prs/ukb-dbp-both.profile") %>%
  select(IID, SCORE) %>%
  rename(DBP_SCORE=SCORE)

covs <- fread("/finngen/red/ht_prs/data/R5_cov_pheno_selected")


```


**Combining data**


```{r}
df <- endpoints %>%
  left_join(sbp_prs, by = c("FINNGENID" = "IID")) %>%
  left_join(dbp_prs, by = c("FINNGENID" = "IID")) %>%
  left_join(covs, by = c("FINNGENID" = "FINNGENID")) 
 
#Removing cases without genetic information and some other preprocessing
df <- df %>%
  filter(!is.na(batch)) %>%  
  mutate_at(c("batch","cohort"),as.factor) 

dim(df)

```


**Properties of variables **


```{r, R.options = list(width = 120) }

df_tmp <- df %>% select(I9_HYPTENS_AGE, I9_HYPTENS, SBP_SCORE, DBP_SCORE)
summary(df_tmp)

df_tmp2 <- df %>% select(I9_CVD_HARD, I9_CVD_HARD_AGE, I9_CHD, I9_CHD_AGE, I9_STR, I9_STR_AGE)
summary(df_tmp2)

dim(df)

#Summary information about sex and age at the end of followup
df$SEX_IMPUTED %>%
  mean()
  
summary(df$FU_END_AGE)
sd(df$FU_END_AGE)


```    


<details><summary>**Visualisations on data**</summary>


```{r, R.options = list(width = 120) }

df_scores <- df %>% select(SBP_SCORE,DBP_SCORE)
cor(df_scores, method = "pearson")

par(mfrow=c(1,2))
hist(df$SBP_SCORE)
hist(df$DBP_SCORE)
par(mfrow=c(1,1))

#5000 random observations
df %>% 
  sample_n(5000) %>%
  vis_dat()

#Visualisation of endpoints

endpoints <- c("I9_HYPTENS", "I9_CVD_HARD", "I9_CHD", "I9_STR")
endpoints_age <- c("I9_HYPTENS_AGE","I9_CVD_HARD_AGE", "I9_CHD_AGE", "I9_STR_AGE")

df_long_ep <-   df %>%
  select("FINNGENID", all_of(endpoints)) %>%
  pivot_longer(cols=endpoints, names_to = "endpoint",values_to = "status")

df_long <-   df %>%
  select("FINNGENID", all_of(endpoints_age)) %>%
  pivot_longer(cols=endpoints_age, names_to = "endpoint",values_to = "age") %>%
  mutate(endpoint = str_remove(endpoint,"_AGE")) %>%
  left_join(df_long_ep, by=c("FINNGENID", "endpoint"))

ggplot(df_long, aes(x=age, color=as.factor(status), fill=as.factor(status))) +
    geom_histogram(alpha=0.4) +
    theme(legend.title = element_blank()) +
    facet_wrap(~endpoint, ncol=2)

#PC:s

df_long_pc <-   df %>%
  select("FINNGENID", PC1:PC10) %>%
  pivot_longer(cols=PC1:PC10, names_to = "PC", values_to = "value")

ggplot(df_long_pc, aes(x=value)) +
    geom_histogram() +
    theme(legend.title = element_blank()) +
    facet_wrap(~PC, ncol=4)


#Other covariates
df_tmp3 <- df %>% select(BL_YEAR, batch, cohort)
summary(df_tmp3)

```

</details>
<br>


**Creating scaled variables**

```{r}
df <- df %>%
  mutate(SBP_SCALED = scale(SBP_SCORE),
         DBP_SCALED = scale(DBP_SCORE))
```



**Dividing to late and early hypertension**


```{r}

df <- df %>%  
  mutate(HT_EARLY = case_when(I9_HYPTENS_AGE < 55 ~ I9_HYPTENS,
                              is.na(I9_HYPTENS) ~ I9_HYPTENS,
                              TRUE ~ as.integer(0)),
         HT_LATE  = case_when(I9_HYPTENS_AGE >= 55 ~ I9_HYPTENS,
                              is.na(I9_HYPTENS) ~ I9_HYPTENS,
                              TRUE ~ as.integer(0)),
         AGE_CAT  = if_else(I9_HYPTENS_AGE <  55, 1,2))

```



**Creating categorical variables**

Data was categorized also in Mars *et al.*, Nat.Med. 26, 549-557, (2020). 


```{r}

#Let's add parametric versions of the scores
my_probs = c(0,0.025,0.2,0.8,0.975,1)
my_labels =  c("<2.5%","2.5-20%","20-80%","80-97.5%",">97.5%")

df <- df %>%  
  mutate(
    SBP_CAT = cut(SBP_SCORE, breaks=quantile(df$SBP_SCORE, probs = my_probs), labels=my_labels, include.lowest = T),
    DBP_CAT = cut(DBP_SCORE, breaks=quantile(df$DBP_SCORE, probs = my_probs), labels=my_labels, include.lowest = T)    
  ) 

#Let's create new dataframe for tables with 20-80% as reference. 

df_r <- df %>%
  mutate(
    SBP_CAT = relevel(SBP_CAT, ref = "20-80%"),
    DBP_CAT = relevel(DBP_CAT, ref = "20-80%")
  )


```



# BP PRSs and hypertension


## Proportional hazard assumption: log log curve


**Model by survfit**


```{r}

#SBP
km.sbp <- survfit(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_CAT, data=df)
#print(km.sbp)

#DBP
km.dbp <- survfit(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ DBP_CAT, data=df)
#print(km.dbp)


```


**log log curve**


```{r, fig.width=10, fig.height=5}

cols= c("skyblue", "blue", "black", "red", "salmon")
labels= c("<2.5%","2.5-20%","20-80%","80-97.5%",">97.5%")

par(mfrow=c(1,2))

plot(km.sbp, fun="cloglog", xlim=c(20,100), main="SBP: log log", xlab="Age", col = cols)
plot(km.dbp, fun="cloglog", xlim=c(20,100), main="DBP: log log", xlab="Age", col = cols)

par(mfrow=c(1,1))

```

These look parallel enough. We can use cox model.


## Cox model: covariate adjusted, continuous score



**Running models**


```{r}
#Fitting cox for SBP_SCORE
cx.sbp.cs <- coxph(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_SCALED + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df)

#Fitting cox for DBP_SCORE
cx.dbp.cs <- coxph(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ DBP_SCALED + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df)

```


**Table: Hazard rations:**


```{r}

my_hr_table_cont2(list(cx.sbp.cs, cx.dbp.cs), c("SBP", "DBP")) %>%
  knitr::kable()

```




<details><summary>**All coefficients**</summary>


```{r}

summary(cx.sbp.cs)
summary(cx.dbp.cs)

```


</details>
<br>




## Cox model: adjusted, continuous, by age



```{r}
#Young
cx.sbp.cs.y <- coxph(Surv(I9_HYPTENS_AGE, HT_EARLY) ~ SBP_SCALED + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df)
cx.dbp.cs.y <- coxph(Surv(I9_HYPTENS_AGE, HT_EARLY) ~ DBP_SCALED + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df)

#Old
cx.sbp.cs.o <- coxph(Surv(I9_HYPTENS_AGE, HT_LATE) ~ SBP_SCALED + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df)
cx.dbp.cs.o <- coxph(Surv(I9_HYPTENS_AGE, HT_LATE) ~ DBP_SCALED + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df)

```


**Hazard ratios/early**


```{r}

my_hr_table_cont2(list(cx.sbp.cs.y, cx.dbp.cs.y), c("SBP", "DBP")) %>%
  knitr::kable()


```


**Hazard ratios/late**

```{r}

my_hr_table_cont2(list(cx.sbp.cs.o, cx.dbp.cs.o), c("SBP", "DBP")) %>%
  knitr::kable() 

```



## Cox model: covariate adjusted, categorized score



**Running models**


```{r}
#Fitting cox for SBP_SCORE
cx.sbp <- coxph(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_CAT + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#Fitting cox for DBP_SCORE
cx.dbp <- coxph(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ DBP_CAT + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

```


**Table: Hazard ratios:**


```{r}

my_hr_table2(cx.sbp, cx.dbp) %>%
  knitr::kable()

```


<details><summary>**All coefficients**</summary>


```{r}

summary(cx.sbp)
summary(cx.dbp)


```


</details>
<br>


**Selecting covariates for plot**

We used **mean** values for all numeric variables. For 'batch', we use the level with **lowest significance**.


<details><summary>Choosing covariate values</summary>


```{r, eval=F}

sbp.coefs <-  summary(cx.sbp)$coefficients %>%
  as.data.frame() %>%
  mutate(name = row.names(.)) %>%
  rename(pval="Pr(>|z|)") %>%
  select("name", "coef", "exp(coef)","se(coef)", "pval")

#Let's look bit more on batches to select a one for the figure
batch.coefs <- sbp.coefs %>% 
  arrange(pval) 

batch.coefs  #Let's print out this
  
#Lowest significance is: AxiomGT1_b07_V2P2.calls

```


</details>
<br>


**Cox-plot adjusted**

Creating covariate combinations for the plot.


```{r}

df.sbp <- my_expand(cx.sbp, "SBP_CAT", df, batch = "AxiomGT1_b06_V2P2.calls")
df.dbp <- my_expand(cx.dpp, "DBP_CAT", df, batch = "AxiomGT1_b06_V2P2.calls")


fit.sbp <- survfit(cx.sbp, newdata = df.sbp)
fit.dbp <- survfit(cx.dbp, newdata = df.dbp)

```

And the plot:

```{r, fig.width=10, fig.height=4}


p1 <- my_ggcoxplot(fit.sbp, df.sbp, "SBP", legend="none", ylab="Cumulative hypertension rate") 
p2 <- my_ggcoxplot(fit.dbp, df.dbp,  "DBP", legend="none", ylab="") 

#Legends as separate figure.
p.tmp <- my_ggcoxplot(fit.sbp, df.sbp, "SBP", legend="right") 
lgnd <- get_legend(p.tmp)   
   
#All collected together
splots <- grid.arrange(p1, p2, lgnd, nrow = 1, widths = c(3,3,1))

ggsave(file = "cx_ht_prs.tiff", plot = splots, height = 3.6, width = 9, dpi = 600)
#ggsave(file = "cx_ht_prs.pdf", plot = splots, height = 4, width = 10, dpi = 150)
ggsave(file = "cx_ht_prs.eps", device = cairo_ps, plot = splots, height = 3.6, width = 9, dpi = 150)

```



## Early and late onset hypertension.


**Running models**


```{r}
#Fitting cox for SBP_SCORE
cx.sbp.y <- coxph(Surv(I9_HYPTENS_AGE, HT_EARLY) ~ SBP_CAT + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)
cx.sbp.o <- coxph(Surv(I9_HYPTENS_AGE, HT_LATE)  ~ SBP_CAT + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#Fitting cox for DBP_SCORE
cx.dbp.y <- coxph(Surv(I9_HYPTENS_AGE, HT_EARLY) ~ DBP_CAT + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)
cx.dbp.o <- coxph(Surv(I9_HYPTENS_AGE, HT_LATE)  ~ DBP_CAT + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)


```


**Table: Hazard ratios/early:**


```{r}

my_hr_table2(cx.sbp.y, cx.dbp.y, "HT_EARLY") %>%
  knitr::kable()

```

**Table: Hazard ratios/late:**


```{r}

my_hr_table2(cx.sbp.o, cx.dbp.o, "HT_LATE") %>%
  knitr::kable()

```



## Restricted mean survival times

Pairwise statistics between reference risk (20-80% quantile) and each other quantile is are calculated. For this purpose we need separate dataframe for each statistic of interest. 


**Creating dataframes for each cases:**
 

```{r}

q <- c("<2.5%", "2.5-20%", "80-97.5%", ">97.5%" )
df.list.sbp  <-  lapply(q, my_extr_cats2, cat="SBP_CAT", df=df)
df.list.dbp  <-  lapply(q, my_extr_cats2, cat="DBP_CAT", df=df)

```


**rmst2 is runned for all cases**

Unadjusted analysis - with covariates R is slow and output of this package does not include mean ages for adjusted.

```{r}

rmst0.sbp <- lapply(df.list.sbp, function(x) rmst2(x$I9_HYPTENS_AGE, x$I9_HYPTENS, x$CAT, tau=95))
rmst0.dbp <- lapply(df.list.dbp, function(x) rmst2(x$I9_HYPTENS_AGE, x$I9_HYPTENS, x$CAT, tau=95))

```


**Results are extracted for the table**


```{r}

#Results extracted and collected to table

rmst.table <- 
  my_rmst_coefs(rmst0.sbp) %>%
  bind_rows(my_rmst_coefs(rmst0.dbp)) %>%
  my_tidy_table("0", 1) %>%
  add_row(q="SBP", est="", pval="", lower_mean=NA, upper_mean=NA, mean=NA,  .before = 1) %>%
  add_row(q="", est="", pval="", lower_mean=NA, upper_mean=NA, mean=NA,  .before = 7) %>%
  add_row(q="DBP", est="", pval="", lower_mean=NA, upper_mean=NA, mean=NA, .before = 8)


#rmst.table divided to two and formatted for forestplot

rmst.means <- rmst.table %>%
  select(mean, lower_mean, upper_mean) %>%
  rename(lower = lower_mean, upper=upper_mean) %>%
  add_row(mean=NA, lower=NA, upper=NA, .before = 1)

rmst.text <- rmst.table %>%
 select(q, est, pval) %>%
  mutate(pval = sub("e(-\\d+)", "%*%10^{\\1}", pval, perl=T)) %>%     #to create proper exponents...
  add_row(q=NA, est="Difference, Years (95% CI)", pval="P-value", .before = 1)

#To create proper exponents... I need actual values as 'expression' and other lines as 'character'.
#list format is needed to accomodate different datatypes; parse can convert character unlike 'as.expression'
pvals <- lapply(rmst.text$pval, function(x) {ifelse(grepl("10",x), parse(text = x), x)})
rmst.text.list <- list(as.list(rmst.text$q), as.list(rmst.text$est), pvals)     
```


**Forestplot**


```{r,fig.width=9, fig.height=5.5}

setEPS()
postscript(file = 'rmst_unadj.eps', width = 9, height = 4.5) 

my_rmst_plot(rmst.text.list, rmst.means, xlab = "Age of hypertension onset",
             clip=c(60,90), xticks = c(60, 65, 70, 75,  80, 85), zero=60)
dev.off()

my_rmst_plot(rmst.text.list, rmst.means, xlab = "Age of hypertension onset",
             clip=c(60,90), xticks = c(60, 65, 70, 75,  80, 85), zero=60)

```


# BP PRSs and cardiovascular outcomes


## Proportional hazard assumption: log log curve:


**Model by survfit**


```{r}

#SBP
km.sbp.cvd <- survfit(Surv(I9_CVD_HARD_AGE, I9_CVD_HARD) ~ SBP_CAT, data=df)
#print(km.sbp.cvd)

#DBP
km.sbp.chd <- survfit(Surv(I9_CHD_AGE, I9_CHD) ~ SBP_CAT, data=df)
#print(km.sbp.chd)

#PP
km.sbp.str <- survfit(Surv(I9_STR_AGE, I9_STR) ~ SBP_CAT, data=df)
#print(km.sbp.str)


```



**log log curve:**


```{r, fig.width=10, fig.height=8}

cols= c("skyblue", "blue", "black", "red", "salmon")
labels= c("<2.5%","2.5-20%","20-80%","80-97.5%",">97.5%")

par(mfrow=c(2,2))

plot(km.sbp.chd, fun="cloglog", xlim=c(40,100), main="SBP CHD: log log", xlab="Age", col = cols)
plot(km.sbp.cvd, fun="cloglog", xlim=c(40,100), main="SBP CVD: log log", xlab="Age", col = cols)
plot(km.sbp.str, fun="cloglog", xlim=c(40,100), main="SBP STR: log log", xlab="Age", col = cols)

par(mfrow=c(1,1))

```

These look parallel enough. We can use cox model.


## Cox model: covariate adjusted, continuous score


**Running models**


```{r}

#Fitting cox for CVD
cx.sbp.cvd.cs <- coxph(Surv(I9_CVD_HARD_AGE, I9_CVD_HARD) ~ SBP_SCALED + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df)

#Fitting cox for CHD
cx.sbp.chd.cs <- coxph(Surv(I9_CHD_AGE, I9_CHD) ~ SBP_SCALED + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df)

#Fitting cox for STR
cx.sbp.str.cs <- coxph(Surv(I9_STR_AGE, I9_STR) ~ SBP_SCALED + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df)

```



**Table: Hazard rations:**


For adjusted cox-model.


```{r}

my_hr_table_cont2(list(cx.sbp.cvd.cs, cx.sbp.chd.cs, cx.sbp.str.cs), c("CVD", "CHD", "Stroke")) %>%
  knitr::kable()

```




<br>

<details><summary>**All coefficients**</summary>


```{r}

summary(cx.sbp.cvd.cs)
summary(cx.sbp.chd.cs)
summary(cx.sbp.str.cs)


```


</details>
<br>



## Cox model: covariate adjusted, categorized score


**Running models**


```{r}
 
#Fitting cox for CVD_HARD vs. SBP_CAT
cx.sbp.cvd <- coxph(Surv(I9_CVD_HARD_AGE, I9_CVD_HARD) ~ SBP_CAT + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#Fitting cox for CHD vs. SBP_CAT
cx.sbp.chd <- coxph(Surv(I9_CHD_AGE, I9_CHD) ~ SBP_CAT + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#Fitting cox for STR vs. SBP_CAT
cx.sbp.str <- coxph(Surv(I9_STR_AGE, I9_STR) ~ SBP_CAT + SEX_IMPUTED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

```


**Table: Hazard rations:**


```{r}

my_hr_table_cvd2(cx.sbp.cvd, cx.sbp.chd, cx.sbp.str, "SBP_CAT", df) %>%
  knitr::kable()


```



<br>

<details><summary>**All coefficients**</summary>


```{r}

summary(cx.sbp.cvd)
summary(cx.sbp.chd)
summary(cx.sbp.str)


```


</details>
<br>


**Selecting covariates for plot**

We used **mean** values for all numeric variables. For 'batch', we use batch with **low significance** for all variables.


<details><summary>Choosing covariate values</summary>


```{r, eval=F}

summary(cx.sbp.cvd)$coefficients %>%
  as.data.frame() %>%
  mutate(name = row.names(.)) %>%
  rename(pval="Pr(>|z|)") %>%
  select("name", "coef", "exp(coef)","se(coef)", "pval") %>%  
  arrange(pval)

summary(cx.sbp.chd)$coefficients %>%
  as.data.frame() %>%
  mutate(name = row.names(.)) %>%
  rename(pval="Pr(>|z|)") %>%
  select("name", "coef", "exp(coef)","se(coef)", "pval")%>%  
  arrange(pval)

summary(cx.sbp.str)$coefficients %>%
  as.data.frame() %>%
  mutate(name = row.names(.)) %>%
  rename(pval="Pr(>|z|)") %>%
  select("name", "coef", "exp(coef)","se(coef)", "pval")%>%  
  arrange(pval)

#Batch:  AxiomGT1_b36_V2.calls, low for all cases. 

```


</details>
<br>



**Cox-plot: covariate adjusted, categorized score**

```{r}

df.sbp.cvd <- my_expand(cx.sbp.cvd, "SBP_CAT", df, batch = "AxiomGT1_b36_V2.calls")
df.sbp.chd <- my_expand(cx.sbp.chd, "SBP_CAT", df, batch = "AxiomGT1_b36_V2.calls")
df.sbp.str <- my_expand(cx.sbp.str, "SBP_CAT", df, batch = "AxiomGT1_b36_V2.calls")

fit.sbp.cvd <- survfit(cx.sbp.cvd, newdata = df.sbp.cvd)
fit.sbp.chd <- survfit(cx.sbp.chd, newdata = df.sbp.chd)
fit.sbp.str <- survfit(cx.sbp.str, newdata = df.sbp.str)

```


And the plot:


```{r, fig.width=10, fig.height=4}


p1 <- my_ggcoxplot(fit.sbp.cvd, df.sbp.cvd, "CVD", conf.int = T, ylim=c(0,0.5), legend="none", vjust=-8)
p2 <- my_ggcoxplot(fit.sbp.chd, df.sbp.chd,  "CHD", conf.int = T, ylim=c(0,0.5), legend="none", vjust=-8, ylab="")
p3 <- my_ggcoxplot(fit.sbp.str, df.sbp.str, "Stroke", conf.int = T, ylim=c(0,0.5), legend="none", vjust=-8, ylab="")

#legends
p.tmp <-  my_ggcoxplot(fit.sbp.cvd, df.sbp.cvd, "CVD", conf.int = T, ylim=c(0,0.5), legend="right") 
lgnd <- get_legend(p.tmp)  

splots <- grid.arrange(p1, p2, p3, lgnd, nrow = 1, widths = c(3,3,3,1))

ggsave(file = "cx_cvd_prs.png", plot = splots, height = 3, width = 10, dpi = 150)
#ggsave(file = "cx_cvd_prs.pdf", plot = splots, height = 3, width = 10, dpi = 150)
ggsave(file = "cx_cvd_prs.eps", device = cairo_ps, plot = splots, height = 3, width = 10, dpi = 150)

```



# Restricted mean survival times


Pairwise statistics between reference risk (20-80% quantile) and each other quantile is are calculated. For this purpose we need separate dataframe for each statistic of interest. 


**Creating dataframes for each cases:**


```{r}

#Run with covariates stalled - only unadjusted analysis is shown.

q <- c("<2.5%", "2.5-20%", "80-97.5%", ">97.5%" )
df.list.cvd  <-  lapply(q, my_extr_cats2, endpoints=c("I9_CVD_HARD_AGE", "I9_CVD_HARD"), cat="SBP_CAT", df=df)
df.list.chd  <-  lapply(q, my_extr_cats2, endpoints=c("I9_CHD_AGE", "I9_CHD"), cat="SBP_CAT", df=df)
df.list.str  <-  lapply(q, my_extr_cats2, endpoints=c("I9_STR_AGE", "I9_STR"), cat="SBP_CAT", df=df)


```


**rmst2 is runned for all cases**

Unadjusted analysis - with covariates R stalls. Propably too large dataset. [The new version of package faster]

```{r}

#Unadjusted, this works:

rmst0.cvd <- lapply(df.list.cvd, function(x) rmst2(x$I9_CVD_HARD_AGE, x$I9_CVD_HARD, x$CAT, tau=95))
rmst0.chd <- lapply(df.list.chd, function(x) rmst2(x$I9_CHD_AGE, x$I9_CHD, x$CAT, tau=95))
rmst0.str  <- lapply(df.list.str, function(x) rmst2(x$I9_STR_AGE, x$I9_STR, x$CAT, tau=95))


```


**Results are extracted for the table**


```{r}

#Results extracted and collected to table

rmst.table.cvd <- 
  my_rmst_coefs(rmst0.cvd) %>%
  bind_rows(my_rmst_coefs(rmst0.chd)) %>%
  bind_rows(my_rmst_coefs(rmst0.str)) %>%
  my_tidy_table("0", 1) %>%
  add_row(q="CVD", est="", pval="", lower_mean=NA, upper_mean=NA, mean=NA,  .before = 1) %>%
  add_row(q="",    est="", pval="", lower_mean=NA, upper_mean=NA, mean=NA,  .before = 7) %>%
  add_row(q="CHD", est="", pval="", lower_mean=NA, upper_mean=NA, mean=NA,  .before = 8) %>%
  add_row(q="",    est="", pval="", lower_mean=NA, upper_mean=NA, mean=NA,   .before = 14)%>%
  add_row(q="Stroke", est="", pval="", lower_mean=NA, upper_mean=NA, mean=NA,   .before = 15)


#rmst.table divided to two and formatted for forestplot

rmst.means.cvd <- rmst.table.cvd %>%
  select(mean, lower_mean, upper_mean) %>%
  rename(lower = lower_mean, upper=upper_mean) %>%
  add_row(mean=NA, lower=NA, upper=NA, .before = 1)

rmst.text.cvd <- rmst.table.cvd %>%
 select(q, est, pval) %>%
  mutate(pval = sub("e(-\\d+)", "%*%10^{\\1}", pval, perl=T)) %>%     #to create proper exponents...
  add_row(q=NA, est="Difference, Years (95% CI)", pval="P-value", .before = 1)

#To create proper exponents... I need actual values as 'expression' and other lines as 'character'.
#list format is needed to accomodate different datatypes; parse can convert character unlike 'as.expression'
pvals <- lapply(rmst.text.cvd$pval, function(x) {ifelse(grepl("10",x), parse(text = x), x)})
rmst.text.list.cvd <- list(as.list(rmst.text.cvd$q), as.list(rmst.text.cvd$est), pvals)     

```


**Forestplot**


```{r,fig.width=9, fig.height=6}

#Forestplot

setEPS()
postscript(file = 'rmst_cvd_unadj.eps', width = 9, height = 6) 

my_rmst_plot(rmst.text.list.cvd, rmst.means.cvd, 
             clip=c(78,91), xticks = c(80, 85, 90), zero=80)
dev.off()

my_rmst_plot(rmst.text.list.cvd, rmst.means.cvd, 
             clip=c(78,91), xticks = c(80, 85, 90), zero=80)

```

